---
title: "2026-04-13_hw07"
author: "Chris"
date: "2026-04-13"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    css: style.css
---


# Read the Data

```{r message=FALSE}
library(tidyverse)
library(lmerTest)

df_polite <- read_csv("data/HW7_politeness_data.csv") |>
  janitor::clean_names() |>
  mutate(
    gender = factor(gender, levels = c("M", "F")),
    subject = factor(subject),
    scenario = factor(scenario),
    attitude = factor(attitude, levels = c("inf", "pol")),
    frequency = as.numeric(frequency)
  )
```


# Question a

## Gender & Pitch

```{r}
fig_1 <- df_polite |>
  ggplot(aes(x = gender, y = frequency)) + geom_boxplot(aes(fill = gender)) +
  labs(
    x = "Gender",
    y = "Pitch",
    title = "Pitch by Gender"
  ) + theme_minimal()

fig_1
ggsave("plots/2026-04-13_fig1-boxplot_gender-pitch.png", fig_1)
```


## Attitude & Pitch

```{r}
fig_2 <- df_polite |>
  ggplot(aes(x = attitude, y = frequency)) + geom_boxplot(aes(fill = attitude)) +
  labs(
    x = "Attitude",
    y = "Pitch",
    title = "Pitch by Attitude"
  ) + theme_minimal()

fig_2
ggsave("plots/2026-04-13_fig2-boxplot_attitude-pitch.png", fig_2)
```




## Pitch by Gender & Attitude

```{r}
fig_3 <- df_polite |>
  ggplot(aes(x = attitude, y = frequency, fill = gender)) + geom_boxplot() +
  labs(
    x = "Attitude",
    y = "Pitch",
    title = "Pitch by Gender and Attitude"
  ) + theme_minimal()

fig_3
ggsave("plots/2026-04-13_fig3-boxplot_gender-attitude-pitch.png", fig_3)
```


# Question b

## Fit the Model

```{r}
fit1 <- lmer(frequency ~ gender + attitude + (1 | subject), data = df_polite)

summary(fit1)
# anova(fit1)
```


## Covariance Matrix for Y_i

```{r}
vc <- VarCorr(fit1)

tau2   <- as.numeric(vc$subject[1, 1])
sigma2 <- sigma(fit1)^2

V_i <- tau2 * matrix(1, 14, 14) + sigma2 * diag(14)
V_i
```

## Covariance Matrix for Fixed Effects

```{r}
vcov(fit1)
```


## Subject-Specific Intercepts

```{r}
ranef(fit1)$subject # random intercept
coef(fit1)$subject # overall intercept + random intercept
```

The BLUPs of the subject-specific random intercepts are the predicted random effects $\hat{b_i}$, obtained from `ranef(fit1)`. If one wants the full subject-specific intercepts, they are $\hat{\beta_0} + \hat{b_i}$, obtainable from `coef(fit1)`

## Residuals

```{r}
resid(fit1)
```





















