---
title: "2026-04-13_hw07"
author: "Chris"
date: "2026-04-13"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    css: style.css
---


# Read the Data

```{r message=FALSE}
library(tidyverse)
library(lmerTest)

df_polite <- read_csv("data/HW7_politeness_data.csv") |>
  janitor::clean_names() |>
  mutate(
    gender = factor(gender, levels = c("M", "F")),
    subject = factor(subject),
    scenario = factor(scenario),
    attitude = factor(attitude, levels = c("inf", "pol")),
    frequency = as.numeric(frequency)
  )
```


# Question a

## Gender & Pitch

```{r}
fig_1 <- df_polite |>
  ggplot(aes(x = gender, y = frequency)) + geom_boxplot(aes(fill = gender)) +
  labs(
    x = "Gender",
    y = "Pitch",
    title = "Pitch by Gender"
  ) + theme_minimal()

fig_1
ggsave("plots/2026-04-13_fig1-boxplot_gender-pitch.png", fig_1)
```


## Attitude & Pitch

```{r}
fig_2 <- df_polite |>
  ggplot(aes(x = attitude, y = frequency)) + geom_boxplot(aes(fill = attitude)) +
  labs(
    x = "Attitude",
    y = "Pitch",
    title = "Pitch by Attitude"
  ) + theme_minimal()

fig_2
ggsave("plots/2026-04-13_fig2-boxplot_attitude-pitch.png", fig_2)
```




## Pitch by Gender & Attitude

```{r}
fig_3 <- df_polite |>
  ggplot(aes(x = attitude, y = frequency, fill = gender)) + geom_boxplot() +
  labs(
    x = "Attitude",
    y = "Pitch",
    title = "Pitch by Gender and Attitude"
  ) + theme_minimal()

fig_3
ggsave("plots/2026-04-13_fig3-boxplot_gender-attitude-pitch.png", fig_3)
```

Based on the data and the combined plot, females have higher median frequency than males, regardless of attitude. Within males, the informal group also has a higher median than the formal group.  


# Question b

## Fit the Model

```{r}
fit1 <- lmer(frequency ~ gender + attitude + (1 | subject), data = df_polite, REML = FALSE)

summary(fit1)
# anova(fit1)
```


## Covariance Matrix for Y_i

```{r}
vc1 <- VarCorr(fit1)

tau2   <- as.numeric(vc1$subject[1, 1])
sigma2 <- sigma(fit1)^2

V_i <- tau2 * matrix(1, 14, 14) + sigma2 * diag(14)
V_i
```

We fit the random-intercept mixed-effects model
\[
Y_{ij} = \beta_0 + \beta_1 \,\text{gender}_i + \beta_2 \,\text{attitude}_{ij} + b_i + \varepsilon_{ij},
\]
where
\[
b_i \sim N(0,\tau^2), \qquad \varepsilon_{ij} \sim N(0,\sigma^2),
\]
and \(b_i\) and \(\varepsilon_{ij}\) are independent.

For subject \(i\), let
\[
Y_i = (Y_{i1}, Y_{i2}, \dots, Y_{i n_i})^T.
\]
Then the model can be written in matrix form as
\[
Y_i = X_i \beta + Z_i b_i + \varepsilon_i,
\]
where \(Z_i = \mathbf{1}_{n_i}\) for the random-intercept model. Therefore,
\[
\mathrm{Var}(Y_i) = Z_i \,\mathrm{Var}(b_i)\, Z_i^T + \mathrm{Var}(\varepsilon_i)
= \tau^2 J_{n_i} + \sigma^2 I_{n_i}.
\]

Since each subject has 14 observations in this dataset,
\[
\mathrm{Var}(Y_i) = \tau^2 J_{14} + \sigma^2 I_{14}.
\]

Equivalently, the diagonal entries are
\[
\mathrm{Var}(Y_{ij}) = \tau^2 + \sigma^2,
\]
and for \(j \neq k\), the off-diagonal entries are
\[
\mathrm{Cov}(Y_{ij}, Y_{ik}) = \tau^2.
\]

The covariance matrix for the estimates of the fixed effects is
\[
\mathrm{Var}(\hat{\beta}) = (X^T V^{-1} X)^{-1},
\]
which is a \(3 \times 3\) matrix corresponding to the intercept, gender, and attitude effects.

The BLUPs for the subject-specific random intercepts are the predicted random effects
\[
\hat{b}_i = \hat{D} Z_i^T \hat{V}_i^{-1} (Y_i - X_i \hat{\beta}),
\]
which can be obtained in R using \texttt{ranef(fit)\$subject}.

The conditional residuals are
\[
\hat{e}_{ij} = Y_{ij} - \hat{\beta}_0 - \hat{\beta}_1 \,\text{gender}_i - \hat{\beta}_2 \,\text{attitude}_{ij} - \hat{b}_i.
\]

## Covariance Matrix for Fixed Effects

```{r}
vcov(fit1)
```


## Subject-Specific Intercepts

```{r}
ranef(fit1)$subject # random intercept
coef(fit1)$subject # fixed intercept + random intercept
```

The BLUPs of the subject-specific random intercepts are the predicted random effects $\hat{b_i}$, obtained from `ranef(fit1)`. If one wants the full subject-specific intercepts, they are $\hat{\beta_0} + \hat{b_i}$, obtainable from `coef(fit1)`

## Residuals

```{r}
resid(fit1)
```


# Question c

## Fit the Model

```{r}
fit2 <- lmer(frequency ~ gender * attitude + (1 | subject), data = df_polite, REML = FALSE)
summary(fit2)
# anova(fit2)
```


## LRT

```{r}
anova(fit1, fit2)
```


We compared the model with main effects only and the model including the `gender $\times$ attitude` interaction using a likelihood ratio test. The test gave $\chi^2 = 1.3855$ with 1 degree of freedom and p = 0.2392 > 0.05. Since the p-value is greater than 0.05, there is no significant evidence that the interaction term is associated with pitch. Therefore, the simpler model without the interaction term is preferred.  


# Question d

The model is $$Y_{ijk} = \beta_0 + \beta_1 gender_i + \beta_2 attitude_{ijk} + b_i + c_j + \epsilon_{ijk}$$.  
$b_i$ is the random intercept of `subject`, $c_j$ is the random intercept of `scenario` and $\epsilon_{ijk}$ is the residual error. We assume that $b_i \sim N(0, \tau_{sub}^2)$, $c_j \sim N(0, \tau_{scen}^2)$, $\epsilon_{ijk} \sim N(0, \sigma^2)$ and these three terms are mutually independent.  


## Fit the Model

```{r}
fit3 <- lme4::lmer(frequency ~ gender + attitude + (1 | subject) + (1 | scenario), data = df_polite, REML = FALSE)
summary(fit3)
# anova(fit3)
```



## Covariance Matrix

```{r}
# extract variance components
vc3 <- VarCorr(fit3)

tau_subj2 <- as.numeric(vc3$subject[1, 1])   
tau_scen2 <- as.numeric(vc3$scenario[1, 1])  
sigma2    <- sigma(fit3)^2  
```


```{r}
# choose one subject
subj_id <- levels(df_polite$subject)[1]
df_i <- subset(df_polite, subject == subj_id)
df_i <- df_i[order(df_i$scenario, df_i$attitude), ]
```


```{r}
# build matrices
n_i <- nrow(df_i)
J <- matrix(1, n_i, n_i)
A <- outer(df_i$scenario, df_i$scenario, "==") * 1
I <- diag(n_i)
```


```{r}
V_i <- tau_subj2 * J + tau_scen2 * A + sigma2 * I

V_i
```


* Under this model, the covariance matrix for $Y_i$ is the same across subjects, provided that each subject has the same number of observations and the same scenario structure. This is because the variance components $\tau_{sub}^2$, $\tau_{scen}^2$ and $\sigma^2$ are common to all subjects, and the indicator matrix for scenario membership is the same for each subject.  

* For a fixed subject i, there are 14 observations in total, corresponding to 7 scenarios and 2 attitude conditions within each scenario. Let $Y_i$ denote the $14 \times 1$ response vector for subject i. Then the covariance matrix for $Y_i$ is $$Var(Y_i) = \tau_{sub}^2 J + \tau_{scen}^2 A_i + \sigma^2 I$$ where J is a $14 \times 14$ matrix of ones, I is a $14 \times 14$ identity matrix and $A_i$ is the indicator matrix with entry 1 if two observations belong to the same scenario and 0 otherwise.   

* More specifically, the covariance structure is: $$Var(Y_{ijk}) = \tau_{sub}^2 + \tau_{scen}^2 + \sigma^2$$ $Cov(Y_{ijk}, Y_{ijb}) = \tau_{sub}^2 + \tau_{scen}^2$ if the subject and scenario are the same but attitude differs, $Cov(Y_{ijk}, Y_{iak}) = \tau_{sub}^2$ if the subject is the same but the scenario differs, and $Cov(Y_{ijk}, Y_{iab}) = \tau_{sub}^2$ if both scenario and attitude differ but the subject is the same.  

* Thus, within the same subject, observations from the same scenario are more strongly correlated because they share both the subject effect and the scenario effect, whereas observations from different scenarios share only the subject effect.  


## Interpretation


The coefficient for the fixed effect term `attitude` represents the average difference in pitch between the polite (`pol`) and informal (`inf`) conditions, after adjusting for gender and accounting for random intercept variation across subjects and scenarios. In this model, the estimated coefficient is -20.002, which means that, on average, the pitch in the polite condition is about 20 units lower than in the informal condition, holding gender constant.  



































